RL-Musician: A Tool for Music Composition with Reinforcement Learning

### 2020-03-09 docs/paper/paper.tex
refac: Name policy properly, do not call it actor model

\begin{abstract}
In this paper, automatic composition of first species counterpoint is framed as an optimization problem close to reinforcement learning. An agent with shallow neural network as an actor model composes polyphonic pieces sequentially one measure (bar) per time step. Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory. Cross-entropy method is used for searching weights of actor model that result in maximum expected score of a piece.

@@

Brief outline of the setup is as follows. There is an environment that keeps a piece in progress and the piece consists of a fixed number of measures. Each measure must be either empty (if it is not filled yet) or containing exactly one note per a melodic line. Initially, only the first measure and the last measure are filled with user-defined values. At a particular time step of an episode, an agent observes vector in which for every pitch there is an exponential moving average over all measures up to the current one (exclusively) of binary indicators whether this pitch is played at corresponding measure. An action of the agent is filling of the current measure, i.e., adding a pitch to each of the melodic lines. When the last measure is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a so called actor model. These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with genetic algorithms or evolutionary strategies \cite{salimans2017evolution}). From optimization perspective, average over several episodes reward is estimated value of fitness function at the weights used.

Obviously, the approach belongs to reinforcement learning, because there is an environment and an agent, but a dataset is absent. Nevertheless, there are some connections with supervised deep learning. From generation of new pieces point of view, actor model returns probabilities of next sonorities (chords) given current state. This is similar to sampling new sequences from a next-step prediction model trained with maximum likelihood method on a dataset of existing sequences.

@@

For each option, agent concatenates observation $v$ with encoding this option vector $u$ of length $(2s + 1)l$. Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them. Concatenation of $v$ and $u$ is passed as input to so called actor model which is a shallow neural network returning a single real-valued score. After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions. An action is drawn from this distribution and the agent takes it.

@@

The list of suggested further improvements is as follows:
\begin{itemize}
	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged. Also downward skips can be penalized in order to have upward skips before climax and downward conjunct motion after it.
	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes. However, a new term can be added to this average. This term must reflect how different created pieces are from each other.
	\item \textbf{Revise feature representation for actor model.} Probably, procedure of next measure generation should be changed. Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation. To make it less raw, some features inspired by music theory may be engineered.
	\item \textbf{Tune hyperparameters.} In the current software implementation, cross-entropy method has five hyperparameters, an agent has two hyperparameters, and evaluation functions have more than ten hyperparameters. As of now, not so many computational experiments are held in order to find their optimum values. 
	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches. Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates. Moments of time when level of tension changes may be rhythmically accented or unaccented. Close but not exact alignment between rhythmical accent and resolution often results in aesthetic and interesting music. Taking into account constraints of smooth voice leading, optimization problem can be set up based on above considerations.
\end{itemize}

>>>>

\begin{abstract}
In this paper, automatic composition of first species counterpoint is framed as a black-box optimization problem close to reinforcement learning. An agent with shallow neural network as a policy composes polyphonic pieces sequentially one measure (bar) per time step. Once a specified number of measures is added, resulting piece is evaluated based on some hand-written rules derived from music theory. Cross-entropy method is used for searching policy weights that result in maximum expected score of a piece.

@@

Brief outline of the setup is as follows. There is an environment that keeps a piece in progress and the piece consists of a fixed number of measures. Each measure must be either empty (if it is not filled yet) or containing exactly one note per a melodic line. Initially, only the first measure and the last measure are filled with user-defined values. At a particular time step of an episode, an agent observes vector in which for every pitch there is an exponential moving average over all measures up to the current one (exclusively) of binary indicators whether this pitch is played at corresponding measure. An action of the agent is filling of the current measure, i.e., adding a pitch to each of the melodic lines. When the last measure is reached, episode is ended and the agent receives reward based on evaluation of created composition. Trainable parameters of the agent are weights of a neural network used as a policy. These parameters are trained with cross-entropy method \cite{rubinstein1997optimization} (alternatively, they can be trained with evolutionary strategies \cite{salimans2017evolution}, genetic algorithms, simulated annealing, or other black-box optimization methods). From optimization perspective, average over several episodes reward is estimated value of fitness function at the weights used.

Obviously, the approach belongs to reinforcement learning, because there is an environment and an agent, but a dataset is absent. Nevertheless, there are some connections with supervised deep learning. From generation of new pieces point of view, a policy returns probabilities of next sonorities (chords) given current state. This is similar to sampling new sequences from a next-step prediction model trained with maximum likelihood method on a dataset of existing sequences.

@@

For each option, agent concatenates observation $v$ with encoding this option vector $u$ of length $(2s + 1)l$. Exactly one element is 1 and others are 0 amongst the first $(2s + 1)$ elements of $u$; position of the non-zero element corresponds to movement of the first line; the same encoding procedure is valid for other blocks of size $(2s + 1)$ and lines corresponding to them. Concatenation of $v$ and $u$ is passed as input to a policy which is a shallow neural network returning a single real-valued score. After these scores are collected for all valid options, Boltzmann softmax is applied to them in order to convert them to probabilities of actions. An action is drawn from this distribution and the agent takes it.

@@

The list of suggested further improvements is as follows:
\begin{itemize}
	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged. Also downward skips can be penalized in order to have upward skips before climax and downward conjunct motion after it.
	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes. However, a new term can be added to this average. This term must quantify how different created pieces are from each other.
	\item \textbf{Revise feature representation for policy.} Probably, procedure of next measure generation should be changed. Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation. To make it less raw, some features inspired by music theory may be engineered.
	\item \textbf{Tune hyperparameters.} In the current software implementation, cross-entropy method has five hyperparameters, an agent has two hyperparameters, and evaluation functions have more than ten hyperparameters. As of now, not so many computational experiments are held in order to find their optimum values. 
	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches. Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates. Moments of time when level of tension changes may be rhythmically accented or unaccented. Close but not exact alignment between rhythmical accent and resolution often results in aesthetic and interesting music. Taking into account constraints of smooth voice leading, optimization problem can be set up based on above considerations.
\end{itemize}
